---
title: "Pattern associator in R"
output: html_notebook
---
This notebook is intended to illustrate classic approaches to pattern association


```{r}
#Load required libraries
library(plotrix) #Adds plotting functions including ability to draw a circle easily

#Read in pre-built functions

make.pa <- dget("make_pa.r") #Function to create and initialize model
plot.pa <- dget("plot_pa.r") #Function to plot pattern associator
hebb.update <- dget("hebb_update.r") #Function to update weights via Hebb rule.
delta.update <- dget("delta_update.r") #Function to update weights via delta rule.

#Read input and output patterns

#Read CSV file into temporary object:
tmp <- read.csv("patterns1.csv", header = T, row.names = 1)

#Pull out input patterns as a matrix:
inputs <- as.matrix(tmp[,grep("in", names(tmp))])

#Pull out output patterns as a matrix:
targets <- as.matrix(tmp[,grep("out", names(tmp))])

#Get rid of temporary object:
rm(tmp)
```

_inputs_ and _targets_ are matrices containing, respectively, the input and output patterns to be associated. In each matrix, a row corresponds to an item and columns correspond to input or output units in the pattern assosciator.
```{r}
head(inputs)
```
Here you can see each input pattern consists of a vector of +1 or -1 values, which you can think of indicating whether the corresponding unit is active (+1) or inactive (-1). Similarly for the target patterns:
```{r}
head(targets)
```
There are 8 output units, and the target values indicate which activation each output unit should adopt when given the corresponding input pattern.

A pattern associator consists of a set of _weights_ from each input unit direct to each output units. Weights indicate the effect that activity in an input unit has on the connected output unit. **Explain more**

```{r}
m <- make.pa(inputs,targets) #Create model
m
```

```{r}
plot.pa(m, inputs[1,], targets[1,])

```

